Recipient plant traits are important determinants of the impacts of heterospecific pollen upon plant reproduction

### 2019-03-15 Manuscript_draft/Manuscript_draft.Rmd
edits legend

---
title: working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect
author:
- name: Jose B. Lanuza
  affilnum: 1
  email: barragansljose@gmail.com

@@

Pollinator sharing can have negative consequences for plant fitness with the arrival of foreign pollen. However, the costs of heterospecific pollen are not yet well understood. We conducted a glasshouse experiment to understand how phylogenetic relatedness and plant traits mediate the impacts of heterospecific pollen transfer. We conducted 1800 crosses by experimentally transferring pollen (50% and 100% ratio) with reciprocal crosses between 10 species belonging to three different families: Brassicaceae, Solanaceae and Convolvulaceae. Seed set was used as proxy of plant fitness. We found that for 65% of the treatments with 50% mix reduced seed set. Moreover, the reduction in seed set was dependent on the degree of relatedness and reproductive traits of the pollen recipient and not the pollen donor. Our results show that certain traits, particularly compatibility system, are critical in understanding the costs of heterospecific pollen.
\vspace{3mm}\hrule

@@

In most ecosystems, plant species normally coexist and share their floral visitors with other species @waser1996. From the plants’ perspective, pollinator sharing can be positive for some plants @carvalheiro2014 or negative for others @pauw2013, depending on the facilitation gradient. An increasing number of visits often correlates with higher chances of fertilization @engel2003. However this is not always the case, among these possible flower visitors there are also nectar robbers and pollen thieves @inouye1980; @magrach2017. Receiving both sufficient quantity and quality deposited on the stigma is thus highly relevant to the pollination success of the plant @aizen2007.

By visiting many plant species, many pollinators are responsible for conspecific pollen loss and the transport of foreign pollen, both of which can have important detrimental effects on species fitness @morales2008;  @ashman2013; @arceo2016. Foreign pollen arrival can play an important role in plant species fitness but outcomes are variable and appear to be context dependent as there is not always a decrease in fitness @morales2008. Some of this variation is likely due to the enormous variability of foreign pollen transferred across systems ranging from 0 to 75 percent. However, most studies report ranges of heterospecific pollen between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012; @ashman2013; @fang2013, yet even these relatively low amounts of heterospecific pollen transferred can decrease fitness greatly @thomson1982. While we now have some understanding of the impacts of heterospecific pollen quantity, we have less understanding of other factors that could be driving the variation in impacts upon fitness.  @ashman2013 postulated the first predictive framework that identifies a need to understand how plant traits might mediate heterospecific pollen effect, whereby mating system and pollen size were predicted to potentially mediate the impact of foreign pollen transfer on plant fitness. This concept is supported by specific case studies, such as @tong2016 that demonstrate an asymmetrical effect in 6 species of *Pedicularis* whereby the pollen of long styled species was able to grow the full length of the style on short styled species but not vice versa. While this suggests that the impacts of heterospecific pollen may differ among pollen donor and recipient, few studies have been conducted to ascertain whether this pattern is in fact a general trend or to identify the extent to which other plant traits are critical to heterospecific pollen impacts.

Plant traits are crucial to understand heterospecific pollen effect but the multifactorial nature of the traits that are involve in the pollen-pistil interaction make difficult to unravel what are the main traits in driving the effect. These traits can be seen from a male perspective of both donor and recipient where pollen size, pollen aperture number and pollen allelopathy are key components to understand the outcome of foreign pollen arrival @murphy1995; @ashman2013. In @ashman2013 small pollen is predicted to cause a greater fitness decrease, although this can be true there are also other possibilities to consider which can obscure a predictive framework like big pollen can clogg small stigmas with fewer pollen grains, bigger stigmas are less likely to be clogged by small pollen grains and bigger pollen can outcompete smaller pollen grains due to faster pollen tube growth rate @williams1990. Moreover, to understand the different mechanical or chemical effects of pollen also the female traits of the pollen recipient have to be considered, from the literature these main traits are: stigma size, style length, number of ovules, incompatibility system and also a structural trait such as flower morphology @montgomery2012; @ashman2013; @tong2016. For example, greater stigmatic area is positively correlated with greater heterospecific pollen deposition @montgomery2012 and therefore possibly with an increase in negative effect. For species that are self-incompatible the barriers towards heterospecific pollen are stronger than self-compatible species @ashman2013. Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Although past research has progress in the understanding of what traits can mediate the effect as we have shown here, there are multiple traits involved and multiple possible scenarios still to be explored empirically for a full understanding of the importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of pollen can occur is important to consider the relatedness of the interacting species. Closely related species are more likely to have similar traits @letten2015. The similarity in traits between closely related species can lead to higher chances of ovule usurpation/abortion @arceo2011 and therefore studies predict and show a greater negative effect of closely related species @ashman2013; @arceo2016. Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species @galen1989; @neiland1999. Despite the fact that far related species are also able to decrease species fitness (REFS). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013). Understanding the role of foreign pollen from distantly related species thus deserves greater attention. The relatedness of foreign pollen gives a first snapshoot of where the pollen competition can occur and therefore could be a proxy of effect. With that said, @arceo2016 is the only work until our knowledge which has proven a greater effect of close related species through a meta-analysis but with low sample sizes and lack of significance. Therefore, there is a need to study the effect of heterospecific pollen of far and close related species at community level beyond single pairwise interactions.

Heterospecific pollen studies in nature have the complexity added of great environmental variability which can lead to confounding interpretations. Moreover, the great diversity of floral structures and small sizes of some of them such as Asteraceae species make tedious or almost impossible the studies of heterospecific pollen on the field. For this reason, we investigated how floral reproductive traits and relatedness mediate the impact of heterospecific pollen by creating an artificial co-flowering community in a glasshouse with 10 species belonging to three different families with heterogeneous reproductive traits. Our study tries to answer the following questions: (i) Do heterospecific pollen reduce seed set? To what extent do (ii) floral reproductive traits and (iii) relatedness, mediate the impacts of heterospecific pollen on seed set.

@@

sp_legend <- readRDS("Data/sp_legend.RData")
p11<- ggplot(sp_legend, aes(Species_1,Cohen_d, size=10)) + theme_bw(base_size=10)+theme(legend.title = element_blank(),plot.margin = unit(c(0,0,0,0.3), "cm"),axis.text.y = element_blank())+ geom_point(size=2,alpha=c(1,1,1,1,1,1,1,1,1,1),aes(color=factor(Family))) +geom_errorbar(size=0.6,alpha=0.5, aes(x = Family, ymin = Lower, ymax = Upper,color=factor(Family)))+scale_color_manual("Family",values=c("#D55E00","#009E73","#E69F00","#0072B2"))+ scale_fill_manual("Family",values=c("#D55E00", "#009E73", "#0072B2","#0072B2"))+
  xlab("") + ylab("Hedges' g") +geom_hline(yintercept=0, linetype="dashed", color = "black")+ggtitle(substitute(paste("J)",italic(' S. melongena')))) 
 #grid.arrange(p1, p5, p3, p4, nrow=2,ncol=2,top="Solanaceae")
grid.newpage()

@@

 grid.draw(arrangeGrob(cbind(ggplotGrob(p1+theme(legend.position="none")),ggplotGrob(pp2+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p3+theme(legend.position="none")),ggplotGrob(p4+theme(legend.position="none")),size="last"),cbind(ggplotGrob(p5+theme(legend.position="none")),ggplotGrob(p6+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p7+theme(legend.position="none")),ggplotGrob(p8+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p9+theme(legend.position="none")),ggplotGrob(p10+theme(legend.position="none")),size="last"),nrow=6))
#mylegend
```

@@

pp1<- ggplot(some_effect_size, aes(Family,Cohen_d, size=10)) + theme_bw(base_size=10)+ geom_point(size=2,aes(color=factor(Family))) +
  geom_linerange(size=0.6,linetype=1,show.legend=FALSE, aes(x = Family, ymin = Lower, 
ymax = Upper,color=factor(Family)))+
  scale_color_manual("Family",values=c("#E69F00", "#0072B2", "#009E73", "#D55E00"))+
  scale_fill_manual("Family",values=c("#E69F00", "#0072B2", "#009E73", "#D55E00"))+
  xlab("") + ylab("") + rotate(ylim = c(-6, 2))+guides(fill=FALSE)+
  geom_hline(yintercept=0, linetype="dashed", color = "black")

@@

mylegend<-g_legend(pp1)

@@

>>>>

---
title: Recipient plant traits are important determinants of the impacts of heterospecific pollen upon plant reproduction
author:
- name: Jose B. Lanuza
  affilnum: 1
  email: barragansljose@gmail.com

@@

Pollinator sharing can have negative consequences for plant fitness with the arrival of foreign pollen. However, the mechanisms underlying the variation in outcomes as a result of pollen contribution by different plant species are not yet well understood. We conducted a glasshouse experiment to understand how plant traits and phylogenetic relatedness mediate the impacts of 15 heterospecific pollen transfer. We conducted 1800 reciprocal crosses by experimentally transferring pollen (50%  and 100% foreign pollen ratio) between 10 species belonging to three different families: Brassicaceae, Solanaceae and Convolvulaceae. Seed set was used as proxy of plant fitness. In the treatments of 100% foreign pollen, we found reduced seed set in X% of the treatments. In the treatments of 50% foreign pollen, we found reduced seed set in 65% of the treatments. Moreover, the reduction in seed set was dependent on the reproductive traits of the pollen recipient, but not the pollen donor or relatedness. Our results show that certain traits of recipient plants, particularly compatibility system, are critical in understanding the costs of heterospecific pollen.

\vspace{3mm}\hrule

@@

In most ecosystems, plant species normally coexist and share their floral visitors with other species @waser1996; @carvalheiro2014. From the plants’ perspective, pollinator sharing can be positive for some plants as an increasing number of visits often correlates with higher chances of fertilization @engel2003. Yet, among these possible flower visitors there are also nectar robbers, pollen thieves @inouye1980; @magrach2017, and inconstant pollinators that transfer foreign pollen from other plants @pauw2013. By visiting many plant species, many pollinators are responsible for conspecific pollen loss and the transport of foreign pollen, both of which can have important detrimental effects on species fitness @morales2008; @ashman2013; @arceo2016. Receiving both sufficient quantity and quality deposited on the stigma is thus highly relevant to the pollination success of the plant @aizen2007. Foreign pollen arrival can play an important role in plant species fitness but outcomes are variable and appear to be context dependent as there is not always a decrease in fitness @morales2008. Some of this variation is likely due to the enormous variability of foreign pollen transferred across systems ranging from 0 to 75 percent. However, most studies report ranges of heterospecific pollen between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012; @ashman2013; @fang2013, yet even these relatively low amounts of heterospecific pollen transferred can decrease fitness greatly @thomson1982.

While we have some understanding of the impacts of heterospecific pollen quantity, we have little knoweldge of the factors that could be driving the variation in pollen quality  upon fitness. Plant traits are crucial to understand heterospecific pollen effect but the multifactorial nature of the traits that are involved in the pollen-pistil interaction make difficult to unravel exactly which traits are driving the effect. @ashman2013 postulated the first predictive framework that identifies a need to understand how plant traits might mediate heterospecific pollen effect, whereby mating system and pollen size were  predicted to potentially mediate the impact of foreign pollen transfer on plant fitness.

The concept of trait driven mechanisms is not new and is supported by system specific studies.  Pollen size, pollen aperture number and pollen allelopathy are thought to be key components in understanding the outcome of foreign pollen arrival @murphy1995; @ashman2013. For example, small pollen is predicted to decrease plant fitness because XXXXXX. Yet, large pollen can outcompete smaller pollen grains due to faster pollen tube growth rate @williams1990. Hence, understanding the different mechanical or chemical effects of pollen requires knowledge of the female traits of the pollen recipient to also be considered @montgomery2012; @ashman2013; @tong2016. For example, greater stigmatic area is positively correlated with greater heterospecific pollen deposition @montgomery2012 and therefore likely to result in an greater negative effect upon plant fitness. Further, species that are self-incompatible are thought to be more resistant to the negative impacts of heterospecific pollen than self-compatible species @ashman2013.

When both donor and recipient traits are considered together, other combinations of traits are also likely to impact plant fitness.  For example, large pollen grains could potentially clog small stigmas with fewer pollen grains, and larger stigmas are less likely to be clogged by small pollen grains. Yet, few studies have considered how effects might differ among donor and recipient species. @tong2016 demonstrate an asymmetrical effect in 6 species of Pedicularis whereby foreign pollen of long styled species was able to grow the full length of the style on short styled species but not vice versa. While this suggests that the impacts of  heterospecific pollen may differ among pollen donor and recipient, few studies have been conducted to ascertain whether this pattern is in fact a general trend or to identify the extent to which other plant traits are critical to heterospecific pollen impacts.

It is challenging to identify general patterns with respect to the mechanisms driving foreign pollen impacts as results are often obscured by the variability within and among species. Closely related species are more likely to have similar  traits @letten2015. The similarity in traits between closely related species can lead tohigher chances of ovule usurpation/abortion @arceo2011 hence, greater negative effects of HP pollen are thought to be a associated with more closely related species @ashman2013; @arceo2016. Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species @galen1989; @neiland1999 and those that have, often report low sample sizes and a lack of significance. Therefore, there is a need to study the effect of heterospecific pollen of far and close related species at community level beyond single pairwise interactions. Given that pollen carried on many insects and stigmas has been found to carry multiple species of foreign pollen with little attention to degree of relatedness @arceo2016; @fang2013. understanding the role of foreign pollen from distantly related species thus deserves greater attention

We investigated how floral reproductive traits and relatedness mediate the impact of heterospecific pollen by creating an artificial co-flowering community in a glasshouse with 10 species belonging to three different families with heterogeneous reproductive traits. Our study addressed the following questions:

1.	To what extent does the amount of foreign pollen applied to stigmas impact plant reproductive fitness (i.e. 50% and 100% foreign pollen ratio.

2.	How do floral reproductive traits and plant relatedness mediate the impacts of heterospecific pollen on seed set.

@@

sp_legend <- readRDS("Data/sp_legend.RData")
p11<- ggplot(sp_legend, aes(Species_1,Cohen_d, size=10)) + theme_bw(base_size=10)+theme(legend.direction="horizontal",legend.title = element_blank(),plot.margin = unit(c(0,0,0,0.3), "cm"),axis.text.y = element_blank())+ geom_point(size=2,alpha=c(1,1,1,1,1,1,1,1,1,1),aes(color=factor(Family))) +geom_errorbar(size=0.6,alpha=0.5, aes(x = Family, ymin = Lower, ymax = Upper,color=factor(Family)))+scale_color_manual("Family",values=c("#D55E00","#009E73","#E69F00","#0072B2"))+ scale_fill_manual("Family",values=c("#D55E00", "#009E73", "#0072B2","#0072B2"))+
  xlab("") + ylab("Hedges' g") +geom_hline(yintercept=0, linetype="dashed", color = "black")+ggtitle(substitute(paste("J)",italic(' S. melongena')))) 
 #grid.arrange(p1, p5, p3, p4, nrow=2,ncol=2,top="Solanaceae")
grid.newpage()

@@

 grid.draw(arrangeGrob(cbind(ggplotGrob(p1+theme(legend.position="none")),ggplotGrob(pp2+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p3+theme(legend.position="none")),ggplotGrob(p4+theme(legend.position="none")),size="last"),cbind(ggplotGrob(p5+theme(legend.position="none")),ggplotGrob(p6+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p7+theme(legend.position="none")),ggplotGrob(p8+theme(legend.position="none")), size="last"),cbind(ggplotGrob(p9+theme(legend.position="none")),ggplotGrob(p10+theme(legend.position="none")),size="last"),mylegend,nrow=6))
#mylegend
```

@@

pp1<- ggplot(some_effect_size, aes(Family,Cohen_d, size=10)) + theme_bw(base_size=10)+ geom_point(size=2,aes(color=factor(Family))) + theme(legend.title = element_blank(),legend.direction="horizontal")+
  geom_linerange(size=0.6,linetype=1, aes(x = Family, ymin = Lower, 
ymax = Upper,color=factor(Family)))+
  scale_color_manual("Family",values=c("#E69F00", "#0072B2", "#009E73", "#D55E00"))+
  scale_fill_manual("Family",values=c("#E69F00", "#0072B2", "#009E73", "#D55E00"))+
  xlab("") + ylab("") + rotate(ylim = c(-6, 2))+guides(fill=FALSE)+
  geom_hline(yintercept=0, linetype="dashed", color = "black")

@@

p11<- ggplot(brol_effect_size, aes(Family,Cohen_d, size=10)) + theme_bw(base_size=10)+theme(legend.direction="horizontal",legend.title = element_blank(),axis.text.y = element_blank())+ geom_point(size=2,aes(color=factor(Family))) +geom_errorbar(size=0.6,alpha=0.5, aes(x = Family, ymin = Lower, ymax = Upper,color=factor(Family)))+scale_color_manual("Family",values=c("#D55E00","#009E73","#E69F00","#0072B2"))
 #grid.arrange(p1, p5, p3, p4, nrow=2,ncol=2,top="Solanaceae")
mylegend<-g_legend(p11)

@@

supporting ideas: 
Species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from Whitehead et al. (2018)).
